---
title: "Introduction"
---
These are the workflow vignettes for the Cyclones package. [Full repositry here.](https://github.com/BBranoff/Cyclones) 
These workflows demonstrate the primary functions of the package, with context as 
to the reasoning behind the methods and the final products.

* [Introduction](#introduction)
* [Get Storms](#get-storm-tabular-data)
* [Get Extents](#make-wind-and-pressure-extent-spatial-features)
* [Get Winds](#build-wind-rasters)
* [Get Rains](#build-precip-rasters)
* [Get Storm Surge](#get-storm-surge)


## Highlights
* Produce raster images of cyclone winds, rains, and storm surge, either at specific times or as aggregate
  * Choose data sources and temporal and spatial resolutions
  * Compare calculated winds against hurricane hunter dropsondes
  
## Future Implementation Ideas
* Integrate NISAR data for:
  * Soil moisture
  * Flooding extent
  * Biomass loss
* Integrate STAR NESDIS SAR for windspeed comparisons
* Integrate LANDSAT for NDVI
* Integrate available LiDAR or NAIP clouds for canopy structure
* Integrate SLOSH for storm surge
* Integrate aspect relative incident wind direction / exposure index

# Introduction
R utilities for producing variable temporal and spatial resolution gridded and time-stacked representations of tropical cyclone wind, precipitation, and storm surge.

A note on lists, rasters, wrapping, and parallelization. Most of the functions in 'Cyclones' produce and operate over lists, primarily because they work nicely with lapply() and its parallel equivalents. This makes it relatively easy to process multiple storms in parallel or to process a single storm in parallel. However, raster data via 'terra' can not be passed across parallel workers unless they are wrapped. Thus, for now, the raster products in Cyclones will sometimes be wrapped and must first be unwrapped before they can be utilized in traditional workflows.  

Parallel processing is mostly only necessary for multiple storms or for high resolution outputs and we recommend users familiarize themselves with the sequential default functions and only then utilize parallel when ready for batch processing. For internal functions, parallelism is done via snowfall at the timestep level. This is best for high resolution (both temporal (\<15 mins) and/or spatial (\<10000 m)) outputs and especially when producing Thin Plate Spline (TPS) outputs, and is probably best utilized if at least 5-10 cpus could be set aside for the effort, but there hasnt been any in-depth test of performance yet. Parallelizing at the storm level is demonstrated below in snowfall, but could be done with other libraries at the user's discretion and also could be done with only a few (\<5) cpus. The functions will not allow you to do both internal and external parallelization, but otherwise, the user is expected to know how many cpus can be safely utilized on their system.

## Installation
Cyclones is no yet available on CRAN but can be downloaded from Github. Probably the easiest way to do this is via the 'remotes' package.

```{r install,eval=FALSE}
install.packages('remotes')
remotes::install_github("BBranoff/Cyclones")
```

# Get Storm Tabular Data
Use the ‘get_storms()’ function to gather the available time series,
location, and meteorological information for a particular storm. If a local, 
pre-downloaded IBTrACS .csv or .nc data source is available, it can be used, 
either as the file location or as a pre-loaded dataset. If not, the data will be
downloaded from the web. 

```{r load-data,echo=FALSE,results='hide',message=FALSE,warning=FALSE}
devtools::load_all()
library(ggplot2)
#allstorms <- get_storms("I:/ibtracs.ALL.list.v04r01.csv")
#allstorms <- stuff$allstorms
#storms <- allstorms[grep("CAMILLE_1969|SINLAKU_2026",names(allstorms))]
#windextents <- lapply(storms,make_extents,t_res=60)
windextents <- readRDS("I:/Cyclone Products/Wind/windextents_60.rds")
```

```{r false-load-data,eval=FALSE}
library(Cyclones)

##  storms can be singular or plural and can be identified specifically, or not
##  HURDAT data are relatively limited compared to IBTrACS and may not have full
##  functionality for wind products, but should be sufficient for precipitation and storm surge.
##  extend timeout if download taking too long
##  options(timeout=300)
storms <- get_storms(source="hurdat",name="Camille",basin="NA",season=1969)
storms <- get_storms(name=c("Maria","IRMA"),basin="NA",season=2017)

##  from local file
storms <- get_storms("./ibtracs.ALL.list.v04r01.csv",name=c("Maria","IRMA"),basin="NA",season=2017)
names(storms)

##  alternatively, all of the data can be downloaded and filtered later
##  we recommend this if wind models need to be constructed in later steps
##  downloading all storms or any distant historical storm will likely 
##  require extending the timeout time
##  options(timeout = 300)
allstorms <- get_storms(ib_filt="ALL")

##  filter entire storm dataset for specific storms
storms <- allstorms[grep("CAMILLE_1969|SINLAKU_2026",names(allstorms))]

## the result is a named list of data pertinent to each storm
## Names are recycled throughout Cyclones and include the name, season, basin, and ID#
names(storms)
```

The 'get_storms()' function will automatically consolidate variables across agencies. If only one value is available for a given variable (for example, among the '_WIND' variable versions of USA_WIND, WMO_WIND, WELLINGTON_WIND etc.), it will be used. But many storms reflect reporting from multiple agencies that need to be consolidated in some way for further processing. Because the 'USA' variables have less missing values than the other agencies across the entire dataset, they are prioritized by default and will be used if other agency values are present. Users can otherwise provide a preference agency (WMO, TOKYO, WELLINGTON etc.) to be considered priority. If the preferred agency is not present, all the other values will be averaged as the consolidated value. However, it is important to note that the USA uses 1 minute intervals to define maximum sustained winds, while most of the rest of the world uses 10 minute intervals. Thus, the consolidation of agencies from different intervals will involve a conversion to the preference interval using a linear relationship based on the full dataset. No consolidation is achieved by setting 'consolidate' to FALSE. Other consolidation preferences can be stipulated by using arguments from the 'cons_stormdat()' function, such as 'pref' = "WMO" to preference the World Meteorological Organization values instead, or by setting 'msw_int' to "10min" to  convert all values to 10 minutes instead. Alternatively, the 'cons_stormdat()' function can be used after 'get_storms(consolidate=FALSE)'. It should be noted that most of Cyclones is currently built to ingest consolidated data, so choosing not to consolidate may impact downstream processing in unpredictable ways. 

# Make Wind And Pressure Extent Spatial Features
In some cases, it may be desired to produce simple spatial features of important storm extents, such as the radius of maximum wind, the eye wall diameter, or the radius of the last closed isobar (ROCI). The make_extents() function will do this utilizing the native data, which is typically in 3hr intervals. The function will also interpolate data to a greater frequency using the 't_res' argument, and can incorporate models via the 'mods' argument to fill in missing extents. See the ['Winds' vignette](https://bbranoff.github.io/Cyclones/Winds.html) for more details. 

```{r false-windextents,eval=FALSE}
##  generate linestring (default) and polygon (optionally) wind extent features for each timestep 
##  in each storm. When used as shown below with multiple storms as an input, the functions will 
##  only perform the action on the first storm
windextents <- make_extents(storms,type=c("linestrings","polygons"))

##  if both geometry types were requested they can be directly accessed
windextents$linestrings
windextents$polygons

##  To apply the function to all storms, use lapply
##  parallel functionality demonstrated in the 'Winds" and 'Rains" vignettes
##  use the t_res parameter to adjust temporal resolution (in minutes)
windextents <- lapply(storms,make_extents,t_res=60)

```

These spatial features can also be good for certain storm visualizations in which a fully gridded raster is not necessary.

```{r plot-extents, echo=FALSE, message=FALSE,warning=FALSE,fig.width=9,fig.height=3.5,fig.cap = "Fig 1A. The results of make_extents() are storm tracks, as well as outer limits of various wind and pressure values. If t_res was utilized, interpolated features are also created to fill in the missing times. Additionally, the 'mods' argument can be used to model wind and pressure values that are missing, which is usually the case for older storms or sometimes for storms outside of USA jurisdiction. In this case, an older storm (Camille) has very little native information and because modeling was not utilized in 'make_extents()', there are no wind or pressure extents. Sinlaku, however, as a modern storm over US interests (Mariana Islands), contains more native data and thus more features to map. Examples of modeled extents are shown in the 'Winds' vignette." }
library(ggplot2)
library(patchwork)

### get world boundaries for maps
countries <- rnaturalearth::ne_countries(scale="Medium",continent = c("north america","south america","Europe","Africa","Oceania"))
Cam <- ggplot(windextents$CAMILLE_1969_NA_1969226N18280|>filter(!grepl("interpolated",source)))+
geom_sf(data=countries|>st_transform(st_crs(windextents[[1]])))+
geom_sf(aes(col=maxwind.ms))+
scale_color_gradientn(
    colours = c("black","lightgreen","#44AA99","#DDCC77","#CC6677","#882255"),
    values = scales::rescale(c(64*0.514, 83*0.514, 96*0.514, 113*0.514,137*0.514)), # breakpoints in data space
    limits = c(0, 80))+
ggtitle("Hurricane Camille - 1969")+
theme_bw()+theme(legend.position="none",plot.title = element_text(vjust = -2))+
coord_sf(crs = sf::st_crs(4326))+
scale_x_continuous(limits=c(-95,-60),n.breaks = 4)+
ylim(15,45)

Sin <- ggplot(windextents$SINLAKU_2026_WP_2026099N09152|>filter(!grepl("interpolated",source)))+
geom_sf(data=countries|>st_transform(st_crs(windextents[[2]])))+
geom_sf(aes(col=kts))+
geom_sf(data=windextents$SINLAKU_2026_WP_2026099N09152|>filter(!grepl("interpolated",source),location=="track points"),aes(col=maxwind.ms/0.514))+
scale_color_gradientn(
    colours = c("black","lightgreen","#44AA99","#DDCC77","#CC6677","#882255"),
    values = scales::rescale(c(64, 83, 96, 113,137)), # breakpoints in data space
    limits = c(0, 150))+
ggtitle("Typhoon Sinlaku - 2026")+
theme_bw()+theme(plot.title = element_text(vjust = -2))+
coord_sf(crs = sf::st_crs(4326))+
  scale_x_continuous(limits=c(130,170),n.breaks=4)+
ylim(0,40)
print(Cam+Sin)
```


```{r plot-extents2, echo=FALSE, fig.width=7,fig.height=7,fig.align='left',fig.cap = "Fig 1B. The results of make_extents() zoomed in, separated by native versus interpolated features, and only showing hurricane force wind features for clarity. Again, notice that Camille has only one native wind extent, which is the radius of maximum wind at landfall. No other wind or pressure extents are provided. Using the 'mods' argument in make_extents() would model missing values. Sinlaku on the other hand, has much more native wind extents." }
Cam <- ggplot(windextents$CAMILLE_1969_NA_1969226N18280|>
                 mutate(source=factor(source,levels=c("native","native & modeled","interpolated","interpolated & modeled"))))+
geom_sf(data=countries|>st_transform(st_crs(windextents[[1]])))+
geom_sf(aes(col=maxwind.ms,lty=source))+
scale_linetype_manual(values=c(1,2))+
scale_color_gradientn(
    colours = c("black","lightgreen","#44AA99","#DDCC77","#CC6677","#882255"),
    values = scales::rescale(c(64*0.514, 83*0.514, 96*0.514, 113*0.514,137*0.514)), # breakpoints in data space
    limits = c(0, 80))+
ggtitle("Hurricane Camille - 1969, Cat. 1+")+
theme_bw()+ theme(legend.position="bottom")+
coord_sf(crs = sf::st_crs(4326))+
xlim(-91,-87)+
ylim(29,31)+
facet_grid(~source)

Sin <- ggplot(windextents[[2]]|>filter(location %in% c("track","track points")|kts>=64)|>
                 mutate(source=factor(source,levels=c("native","native & modeled","interpolated","interpolated & modeled"))))+
geom_sf(data=countries|>st_transform(st_crs(windextents[[2]])))+
geom_sf(aes(col=maxwind.ms,lty=source))+
#geom_sf(data=windextents$SINLAKU_2026_WP_2026099N09152|>filter(!grepl("interpolated",source),location=="track points"),aes(col=maxwind.ms))+
scale_linetype_manual(values=c(1,2))+
scale_linewidth_manual(values=c(.5,.5,1,1))+
scale_color_gradientn(
    colours = c("black","lightgreen","#44AA99","#DDCC77","#CC6677","#882255"),
    values = scales::rescale(c(64*0.514, 83*0.514, 96*0.514, 113*0.514,137*0.514)), # breakpoints in data space
    limits = c(0, 80))+
ggtitle("Typhoon Sinlaku, 2026 - Cat. 1+")+
theme_bw()+
theme(legend.position="none")+
coord_sf(crs = sf::st_crs(4326))+
  scale_x_continuous(limits=c(143.5,147),n.breaks=4)+
ylim(13,17)+
facet_grid(~source)
print(Cam/Sin)
```


# Build Wind Rasters
Next, we interpolate and/or model the winds across space and time, creating a more complete gridded representation that can be useful in any number of applications. This is accomplished through either one of three theoretical models, or through a more empirical Thin Plate Spline, which utilizes the above linestrings. The 'get_wind' function accepts a spatial resolution specification and will interpolate results to the temporal resolution of the input data. Although accuracy and smoothness are optimized at smaller resolutions, computation times suffer. For initial testing, a spatial resolution of 20000 m and a temporal resolution of 60 mins is reasonable and default. All storm layers are projected onto a Lambert Conformal Conic reference system that is centered on the storm's extent, which preserves area across large geographic regions and avoids geographic coordinate issues at the 180th meridian.

'get_winds()' returns raster stacks of wind speed, wind direction, and total potential wind power for each time step in the storm's lifetime. A more detailed workflow is provided in the [Winds vignette](https://bbranoff.github.io/Cyclones/Winds.html). 

```{r false-getwinds,eval=FALSE}
###  the following will report multiple warnings because the Camille storm does not have any complete wind fields
###  which are required for most of the different methods. Willoughby includes a theoretical calculation for the
###  radius of maximum winds, so it is the only method that can be used.
###  see the 'Winds" vignette for more information on including modeled values for these methods
winds <- lapply(windextents,get_wind,methods=c("all"))

###  they are now stored in separate lists, one for each method
winds_Camille <- winds$CAMILLE_1969_NA_1969226N18280
winds_Sinlaku <- winds$SINLAKU_2026_WP_2026099N09152

##  if previously saved to disk at ~./Cyclone Products/Wind/, can also load using the storm directory name
##  if we want the list to have the same names, we have to set them before running as name = directory
winds <- list(CAMILLE_1969_NA_1969226N18280="CAMILLE1969_20000_60_WIND",
              SINLAKU_2026_WP_2026099N09152="SINLAKU2026_20000_60_WIND")
winds <- lapply(winds,get_wind,todir="./Cyclone Products/Wind/",loadrasts = TRUE)

###  to plot specific dates via ggplot
Camdate = c("196908171100","196908180400","196908181200")
###  grab only the wind speed variable (msw)
windsCam_msw <- winds_Camille$willoughby[[grep("msw",names(winds_Camille$willoughby))]]
windsCam_msw <- windsCam_msw[[time(windsCam_msw) %in% as.POSIXct(Camdate,format="%Y%m%d%H%M",tz="UTC")]]
## convert to xyz dataframe
windsCammsw_df <- as.data.frame(project(windsCam_msw,"epsg:4326"),xy=TRUE)
names(windsCammsw_df)[3:5] <- paste(names(windsCammsw_df)[3:5],rep(unique(time(windsCam_msw))),sep="_")
windsCammsw_df <- windsCammsw_df |>
  tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","time"))|>
  mutate(layer=paste0("Camille\n",method))
ggplot(data=windsCammsw_df) +
  geom_raster(aes(x=x,y=y,fill=value))+
  facet_grid(layer~time)+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Maximum\nSustained\nWinds\n(m/s)",
                       breaks=c(0,10,20,30,40,50,60,70))+
  theme_bw()+
  geom_sf(data=rnaturalearth::ne_countries(scale="Medium",country = c("united states of america","mexico")),fill=NA,col="black")+
  ggtitle("Wind Method Comparisons")+
  scale_x_continuous(limits=c(-94.5,-83),breaks=c(-92,-88,-84))+
  ylim(23,36.5)+
  theme(axis.title = element_blank(),panel.grid=element_blank())+
  theme(legend.position = "none")
```

```{r plotwinds1,echo=FALSE,message=FALSE,warning=FALSE,results='hide',fig.width=9,fig.height=11,fig.align='left',fig.cap="Fig 2. The results of get_wind() are rasters for each timestep and at the desired spatial resolution (default in this case is 20km). This is showing windspeed, but wind direction and power dissipation are also provided. For Camille, because much of the required information was missing, only the Willoughby method could be utilized. Sinlaku, however, contains more data and all methods were calculated. Notice, however, the inflated eye region for the TPS method. Although Sinlaku does contain the necessary data for TPS calculation, it is still missing many of the extents when compard to North Atlantic storms. Thus, without using the 'mods' argument in make_extents(), TPS values may not accurately represent the wind extents. A more in-depth exploration of this is given in the 'Winds' vignette."}
library(terra)
library(sf)
winds_Camille <- get_wind("CAMILLE1969_20000_60_WIND",todir="I:/Cyclone Products/Wind/v3_unmodeled/",loadrasts = TRUE)
Camdate = c("196908171100","196908180400","196908181200")
windsCam_msw <- winds_Camille$willoughby[[grep("msw",names(winds_Camille$willoughby))]]
windsCam_msw <- windsCam_msw[[time(windsCam_msw) %in% as.POSIXct(Camdate,format="%Y%m%d%H%M",tz="UTC")]]
windsCammsw_df <- as.data.frame(project(windsCam_msw,"epsg:4326"),xy=TRUE)
names(windsCammsw_df)[3:5] <- paste(names(windsCammsw_df)[3:5],rep(unique(time(windsCam_msw))),sep="_")
windsCammsw_df <- windsCammsw_df |>
  tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","time"))|>
  mutate(layer=paste0("Camille\n",method))
gg1 <- ggplot(data=windsCammsw_df) +
  geom_raster(aes(x=x,y=y,fill=value))+
  facet_grid(layer~time)+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Maximum\nSustained\nWinds\n(m/s)",
                       breaks=c(0,10,20,30,40,50,60,70))+
  theme_bw()+
  geom_sf(data=rnaturalearth::ne_countries(scale="Medium",country = c("united states of america","mexico")),fill=NA,col="black")+
  ggtitle("Wind Method Comparisons")+
  scale_x_continuous(limits=c(-94.5,-83),breaks=c(-92,-88,-84))+
  ylim(23,36.5)+
  theme(axis.title = element_blank(),panel.grid=element_blank())+
  theme(legend.position = "none")

winds_Sinlaku <- get_wind("SINLAKU2026_20000_60_WIND",todir="I:/Cyclone Products/Wind/v3_unmodeled/",loadrasts=TRUE)
Sindate = c("202604141000","202604141600","202604150000")
windsSinmsw_df  <- lapply(winds_Sinlaku, function(x){
  x <- x[[grep("msw",names(x))]]
  x <- x[[time(x) %in% as.POSIXct(Sindate,format="%Y%m%d%H%M",tz="UTC")]]
  df <- as.data.frame(project(x,"epsg:4326"),xy=TRUE)
  names(df)[3:5] <- paste(names(df)[3:5],rep(unique(time(x))),sep="_")
  df |>
  tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","time"))
} )
windsSinmsw_df <- do.call(rbind,windsSinmsw_df)|>
   mutate(layer=paste0("Sinlaku\n",method))
#winds_df <- bind_rows(windsSinmsw_df,windsCammsw_df)
gg2 <-ggplot(data=windsSinmsw_df) +
  geom_raster(aes(x=x,y=y,fill=value))+
  facet_grid(layer~time)+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Maximum\nSustained\nWinds\n(m/s)",
                       breaks=c(0,10,20,30,40,50,60,70))+
  theme_bw()+
  geom_sf(data=rnaturalearth::ne_countries(scale=10,continent="Oceania"),fill=NA,col="black")+
  #ggtitle("Typhoon Sinlaku")+
  scale_x_continuous(limits=c(138,152.5),breaks=c(140,145,150))+ylim(8,22)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position = "bottom")
print((gg1/gg2)+plot_layout(heights = c(1,2)))
```

It may also be desired to obtain an aggregate summary of a storm, rather than or in addition to individual time steps.This can be achieved via the terra package's 'app' functionality. In this case, we simply take the maximum to determine the maximum sustained wind at any given point during the duration of the storm, or the sum to get the total power dissipation at any given point. 

```{r plotwinds_agg,echo=FALSE,message=FALSE,warning=FALSE,results='hide',fig.width=7,fig.height=6,fig.align='left',fig.cap="Fig 3. Aggregating the individual time step layers from get_wind() produces a whole-storm perspective. In this case, the maximum wind is shown. Aggregating can be done in many complicated ways (e.g. the duration of hurricane force winds, the direction of maximum winds etc.) and a dedicated agg_winds() function is forthcoming."}

### take only the msw (maximum sustained winds) layers 
Cam_msw <- app(winds_Camille$willoughby[[grepl("msw",names(winds_Camille$willoughby))]],fun="max",na.rm=TRUE)
terra::plot(Cam_msw,main="Hurricane Camille, 1969 - Maximum Sustained Winds (m/s)")
terra::plot(st_geometry(st_transform(rnaturalearth::ne_countries(),crs(Cam_msw))),add=TRUE)

##  additionally, we could take the sum of the total power dissipation
##  Power was not available for Camille because no pressure field was available (without modeling)
##  here, because there were multiple methods, we add a grouping step and we 'tapp()' instead of 'app()'
#Sin_power <- winds_Sinlaku[[grepl("power",names(winds_Sinlaku))]]
#methods <- unlist(sapply(strsplit(names(Sin_power),"_"),"[[",3))
#Sin_power <- terra::tapp(Sin_power,index=methods,fun="sum",na.rm=TRUE)
#terra::plot(Sin_power)
#mtext("Power dissipation for Super Typhoon Sinlaku, 2026", side=3, line=-2, outer=TRUE, cex=1.5)
```


# Build Precip Rasters
Cyclone precipitation products are different from wind products in two ways: 1) They are mostly retrieved rather than calculated and assembled, and 2) they include 'predecessor' events that occur before the storm's arrival. 

Precipitation can come from one of three sources: The default European Centre for Medium-Range Weather Forecasts (ecmwf), NASA's Global Precipitation Measurement Mission (gpm), or the Multi-Source Weighted-Ensemble Precipitation product (mswep). For all sources, some sort of pre-registration and subsequent configuration in R is necessary to access data. Additional packages are also necessary for API interaction at each source. These will be prompted for when required. Alternatively, data could be downloaded manually and held in the './~dpath~/~source~/' directory, ensuring original filenames are retained for loading via 'get_precip()'. 

* [register with ECMWF for ERA5.](https://www.ecmwf.int/user/login)
* [register with NASA Earth Data for GPM IMERG.](https://urs.earthdata.nasa.gov/)
* [request access to MSWEP](https://www.gloh2o.org/)

Precipitation products also come with a 'prestorm' option, which allows for the collection of precipitation over certain time periods, typically 24-48 hrs, before the storm's arrival. This is to account for the well-document PRE events and is explained in more detail in the ['Rains' vignette](https://bbranoff.github.io/Cyclones/Rains.html). 

```{r get_precip_false,eval=FALSE}
##  to download the ERA5 website, you must first register with either ECMWF or NASA, or request access to MSWEP:
##  [register with ECMWF for ERA5.](https://www.ecmwf.int/user/login)
##  [register with NASA Earth Data for GPM IMERG.](https://urs.earthdata.nasa.gov/)
##  [request access to MSWEP](https://www.gloh2o.org/)

##  once registered with ECMWF, you will recieve an API token that looks like this:  abcd1234-foo-bar-98765431-XXXXXXXXXX
##  This will give you access to the data repositories.
##  To set your api key in R and make it accessible for future use this:
##  (This only needs to be run once per R installation and setup and should not be run every time) 
library(ecmwfr)
wf_set_key(key = "abcd1234-foo-bar-98765431-XXXXXXXXXX")

##  or, once registered at nasa earthdata, run the following with your username and password 
library(earthdatalogin)
edl_netrc(username = "XXXX",pasword="XXXX")

##  or, once the MSWEP data has been granted, access is through Google Drive and can be granted via (follow the pop-up prompts):
library(googledrive)
drive_auth()

##  The system is now configured to access the data.
##  To not have to re-download storm data again, store the files in a local repository (dpath)
##  otherwise, if dpath is not provided, they will be stored in a temporary directory (and 
##  re-downloaded in subsequent R sessions). Either way, the function will 
##  provide the resulting rasters, which can be processed and stored as desired.
##  The default source is ecmwf
googledrive::drive_auth() ## for mswep
precip <- lapply(storms,get_precip,dpath="./Precipitation",sources="all")
###  can also load from disk if already saved
precip <- list(CAMILLE_1969_NA_1969226N18280="CAMILLE1969_25000_60_PRECIP",
               ###  because Sinlaku has different sources and s_res was not specified, they are stored in different directories
              SINLAKU_2026_WP_2026099N09152=c("SINLAKU2026_25000_60_PRECIP",
                                              "SINLAKU2026_5000_60_PRECIP",
                                              "SINLAKU2026_10000_60_PRECIP"))
precip <- lapply(precip,get_precip,todir="./Cyclone Products/Precip/",loadrasts = TRUE)

precip
###  results are broken down by storm, source, and also storm and pre-storm (if desired)
###  MSWEP or GPM were not available for Camille, but all three are available for Sinlaku
###  separate storms and methods
precip_Camille <- precip$CAMILLE_1969_NA_1969226N18280$ecmwf$storm
##  not as simple for Sinlaku because the mswep resolution is different than the ecmwf, so must keep them separate
##  more on forcing the same resolution in the 'Rains' vignette
precip_Sinlaku <- list(precip$SINLAKU_2026_WP_2026099N09152$ecmwf$storm,
                       precip$SINLAKU_2026_WP_2026099N09152$mswep$storm)

##  plotting random times in base plot
plot(precip_Camille[[100:102]])
##  ggplot gives more funcitonality
PrecipsCamille_df <- as.data.frame(project(precip_Camille[[100:102]],'epsg:4326'),xy=TRUE)
names(PrecipsCamille_df)[-c(1,2)] <- paste(names(PrecipsCamille_df)[-c(1,2)],time(precip_Camille[[100:102]]),sep="_")
PrecipsCamille_df <-PrecipsCamille_df |>
   tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","source","time"))
ggplot(data=PrecipsCamille_df) +
   geom_raster(aes(x=x,y=y,fill=value))+
  facet_grid(method~time)+
   scale_fill_viridis_c(n=10,na.value = "transparent",name="Rainfall\n(mm/hr)")+
   coord_sf(crs ="+proj=robin +lon_0=180")+
   theme_bw()+
   geom_sf(data=rnaturalearth::ne_countries(),fill=NA,col="black")+
   ggtitle("Hurricane Camille, Precipitation")+
   scale_x_continuous(limits=c(-94.5,-83),breaks=c(-92,-88,-84))+
   ylim(23,36.5)+
   theme(axis.title = element_blank(),panel.grid=element_blank())

###  same code for Sinlaku not shown
```
```{r plot_precip,echo=FALSE,message=FALSE,warning=FALSE,fig.width=9,fig.height=9,fig.cap="Fig 4. Results of get_precip() are time series rasters of precipitation in the desired spatial resolution. In this case, no resolution was requested, so the native resolution is returned. MSWEP data is only available after 1979 and GPM data fter 1998."}
precip <- list(CAMILLE_1969_NA_1969226N18280="CAMILLE1969_25000_60_PRECIP",
               ###  because Sinlaku has different sources and s_res was not specified, they are stored in different directories
              SINLAKU_2026_WP_2026099N09152=c("SINLAKU2026_25000_60_PRECIP",
                                              "SINLAKU2026_5000_60_PRECIP",
                                              "SINLAKU2026_10000_60_PRECIP"))
precip <- lapply(precip,get_precip,todir="I:/Cyclone Products/Precip/v2",loadrasts = TRUE)
precip_Camille <- precip$CAMILLE_1969_NA_1969226N18280$ecmwf$storm#precip$CAMILLE_1969_NA_1969226N18280$CAMILLE_1969_NA_1969226N18280_precip_ecmwf$storm
precip_Sinlaku <- list(precip$SINLAKU_2026_WP_2026099N09152$ecmwf$storm,#precip$SINLAKU_2026_WP_2026099N09152$SINLAKU_2026_WP_2026099N09152_precip_ecmwf$storm,
                       precip$SINLAKU_2026_WP_2026099N09152$mswep$storm,
                       precip$SINLAKU_2026_WP_2026099N09152$gpm$storm)#precip$SINLAKU_2026_WP_2026099N09152$SINLAKU_2026_WP_2026099N09152_precip_mswep$storm)
PrecipsCamille_df <- as.data.frame(project(precip_Camille[[100:102]],'epsg:4326'),xy=TRUE)
names(PrecipsCamille_df)[-c(1,2)] <- paste(names(PrecipsCamille_df)[-c(1,2)],time(precip_Camille[[100:102]]),sep="_")
PrecipsCamille_df <-PrecipsCamille_df |>
   tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","source","time"))
PrecipsSinlaku_df1 <- as.data.frame(project(precip_Sinlaku[[1]][[time(precip_Sinlaku[[1]]) %in% as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC")]],'epsg:4326'),xy=TRUE)
names(PrecipsSinlaku_df1)[-c(1,2)] <- paste(names(PrecipsSinlaku_df1)[-c(1,2)],as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC"),sep="_")
PrecipsSinlaku_df2 <- as.data.frame(project(precip_Sinlaku[[2]][[time(precip_Sinlaku[[2]]) %in% as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC")]],'epsg:4326'),xy=TRUE)
names(PrecipsSinlaku_df2)[-c(1,2)] <- paste(names(PrecipsSinlaku_df2)[-c(1,2)],as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC"),sep="_")
PrecipsSinlaku_df3 <- as.data.frame(project(precip_Sinlaku[[3]][[time(precip_Sinlaku[[3]]) %in% as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC")]],'epsg:4326'),xy=TRUE)
names(PrecipsSinlaku_df3)[-c(1,2)] <- paste(names(PrecipsSinlaku_df3)[-c(1,2)],as.POSIXct(c("2026-04-12 12:00","2026-04-12 13:00","2026-04-12 14:00"),tz="UTC"),sep="_")

PrecipsSinlaku_df <-bind_rows(PrecipsSinlaku_df1 |>
   tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","source","time")),
   PrecipsSinlaku_df2 |>
   tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","source","time")),
   PrecipsSinlaku_df3 |>
   tidyr::pivot_longer(cols=-c(x,y),names_sep = "_",names_to = c("method","unit","source","time")))
Cam <- ggplot(data=PrecipsCamille_df) +
   geom_raster(aes(x=x,y=y,fill=value))+
   facet_grid(method~time)+
   scale_fill_viridis_c(n=10,na.value = "transparent",name="Rainfall\n(mm/hr)",
                        breaks=c(0,5,10,15,20))+
   theme_bw()+
   geom_sf(data=rnaturalearth::ne_countries(country = "united states of america",scale=10),fill=NA,col="black")+
   ggtitle("Hurricane Camille, Precipitation")+
   scale_x_continuous(limits=c(-94.5,-84),breaks=c(-92,-88,-84))+
   ylim(24,35)+
   theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="right")
Sin <- ggplot(data=PrecipsSinlaku_df) +
   geom_raster(aes(x=x,y=y,fill=value))+
   facet_grid(method~time)+
   scale_fill_viridis_c(n=10,na.value = "transparent",name="Rainfall\n(mm/hr)",
                        breaks=c(0,10,20,30,40,50,60))+
   theme_bw()+
   geom_sf(data=rnaturalearth::ne_countries(scale=10,continent="Oceania"),fill=NA)+
   ggtitle("Super Typhoon Sinlaku, Precipitation")+
  scale_x_continuous(limits=c(145,155),breaks=c(142,148,152))+ylim(6,16)+
   theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="right")
print((Cam/Sin)+plot_layout(heights = c(1, 2)))
```

The above demonstrates the differences between the precipitation sources. GPM for Sinlaku is a recent file (as of July 2026), which has not been verified by gauge data. This verification and subsequent correction happens at least 4-6 months after the event. Thus, users should use caution when urilizing GPM for relatively recent storms. Validation routines are needed to help bound precipitation and test for best approximation, such as those from the drop sondes used to validate wind products. 

Just as for the wind products, precipitation can also be aggregated to provide a summary of the entire storm. Of particular interest for rain is the total accumulation, and perhaps the mean rainfall rate. Although not shown here, pre-storm precipitation associated with potential 'precursor rain events' might also be included in such an aggregation. See the [Rains](https://bbranoff.github.io/Cyclones/Rains.html) vignette for more details. 

```{r plotrain_agg,echo=TRUE,message=FALSE,warning=FALSE,results='hide',fig.width=7,fig.height=7,fig.align='left',fig.cap="Fig 5. Aggregating the individual time step layers from get_precip() produces a whole-storm perspective. In this case, the total accumulated rainfall. Aggregating can also include prestorm precipitation if available."}

### use the 'app()' function from terra to sum across the timesteps
Cam_precip <- app(precip_Camille,fun="sum",na.rm=TRUE)
terra::plot(Cam_precip,main="Hurricane Camille, 1969 - Total Rainfall (mm)")
terra::plot(st_geometry(st_transform(rnaturalearth::ne_countries(),crs(Cam_precip))),add=TRUE)
```

# Get Storm Surge

Currently, data for storm surge is the least complete relative to the other products of wind and precipitation. This is mostly due to a sparsity of global storm surge data. The Cyclones method involves comparing available water level data with elevation data. This assumes surge waters flow evenly around geographic features, which is overly simplistic. Future methods will attempt to incorporate complex coastal hydrology, such as those implemented in the SLOSH models.

Around the United States and it's territories, and especially for more recent storms, there are farm more water level monitoring locations available for the analysis when compared to other locations and more historic storms. When US monitoring data are unavailable, they are supplemented with the Coupled Model Intercomparison Project Phase 6 (CMIP6) data. Although CMIP6 data is available globally and for more historic storms, these modeled water levels are known to under estimate storm surge, which should be considered when evaluating surge products.

The get_surge() function will provide a raster of maximum surge water levels above a given elevation, as well as a simple spatial features object of the water level observations that were incorporated into the analysis. The datum for flooding is the Mean High High Tide or Mean High High Water (MHHW), so surge values are only returned if water levels are higher than this value at any given point.

Unlike the wind and precipitation methods, get_surge() does not allow for spatial or temporal resolution arguments. This is primarily because the elevation data is already high resolution (30 m), and because water depths are not easily generalized across large areas of different elevation, the resulting surge product is also high resolution and very expansive when compared to wind and precipitation. Thus, resulting rasters are relatively large and could easily become obtrusive on even high-performance machines if higher spatial or temporal resolution were requested. For now, get_surge() products are limited to one layer only of the maximum surge depth, and at a moderate spatial resolution of 10 km. 


```{r get_surge_false,eval=FALSE}
###  as with precipitation, surge requires some pre-registration before remote access to the data

###  Elevation comes from the NASA Shuttle Radar Topography Mission
###  it is obtained through the same earthdatalogin methods as the Global Precipitation Mission
###  so if that was already setup, it should be ready
###  if not, see earthdatalogin for permanent registration and setup
###  unlike water levels, elevation data is constant for a given area, regardless of the storm time
###  so, once data for a location has been downloaded (and saved utilizing the dpath arg) it can be
###  directly accessed without downloading for future analyses over the same area

###  likewise, CMIP6 data is obtained through the European Centre for Medium-Range Weather Forecasts (ecmwf)
###  so, if that was already setup for the precipitation data, it will serve here as well
###  if not, see ecmwfr::wf_get_key() and ecmwfr::wf_set_key() for setup
###  for storms after 2024, CMIP6 does not offer daily maximum water levels, only 10 min interval levels
###  so, these downloads will take considerably longer

###  NOAA water levels are retrieved through an anonymous api, so no pre-registration is required
###  more info here: https://api.tidesandcurrents.noaa.gov/api/prod/
###  It should be noted that the server does implement throttling
###  Cyclones methods doe introduce sleep statements to prevent throttling, but users may still encounter issues 
###  when requesting large amounts of data

###  Finally, USGS water level data is also retreived though a semi-anonymous API
###  Users may introduce their own api key, which can be obtained here: https://api.waterdata.usgs.gov/signup/
###  This is not necessary, however, as data will still be retreived without a key
###  but it may boost performance

### as with the other functions, dpath will save component data in a directory for direct future access (no download necessary)
surge_Camille <- get_surge(storms$CAMILLE_1969_NA_1969226N18280,dpath="./Storm Surge/")

ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge level\nabove MHHW\n(m)")+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf_text(data=st_transform(rnaturalearth::ne_countries(type="tiny_countries"),crs(windsSinmsw)),
  #             aes(label = admin), size = 3, check_overlap = TRUE) +
  ggtitle("Hurricane Camille Storm Surge at Louisiana")+
  xlim(-93,-83)+ylim(28,32)+
  theme(axis.title = element_blank(),panel.grid=element_blank())

```

```{r plot_surge,echo=FALSE,message=FALSE,warning=FALSE,fig.width=7,fig.height=10,fig.cap="Fig 6. Results of get_surge() are a single layer only, representing the maximum surge depth above ground for each location during the storm, as well as a corresponding summary table of water level observations. Surge depth is a simple difference between an interpolation of nearby water level observations and the ground elevation. Only depths greater than the mean high high tide are returned. For storms outside of United States jurisdictions, and for any storm before 2000, water level observations are scarce and the CMIP6 model is usually used instead. This model is known to underestimate extreme surge values. Water level sources can be viewed via the corresponding table produced for each storm (below)."}
surge_Camille <- list(rast=rast("I:/Cyclone Products/Storm Surge/v2/CAMILLE_1969_NA_1969226N18280_surge.tif"),points=read.csv("I:/Cyclone Products/Storm Surge/v2/CAMILLE_1969_NA_1969226N18280_surgepoints.csv"))
#surge_Sinlaku <- list(rast=rast("I:/Cyclone Products/Storm Surge/v2/SINLAKU_2026_WP_2026099N09152_surge.tif"),points=read.csv("I:/Cyclone Products/Storm #Surge/v2/SINLAKU_2026_WP_2026099N09152_surgepoints.csv"))

gg1 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge level\nabove MHHW\n(m)")+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  ggtitle("Hurricane Camille Storm Surge at Louisiana")+
  xlim(-93,-83)+ylim(28,32)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
gg2 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge depth\n(m)",
                       breaks=c(0,0.5,1,1.5,2))+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  xlim(-90,-87)+ylim(29.5,31)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
gg3 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge depth\n(m)",
                       breaks=c(0,0.5,1,1.5,2))+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  xlim(-89.5,-89)+ylim(29.8,30.2)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="bottom")
print(gg1/gg2/gg3)
```
```{r surge_table,echo=FALSE,message=FALSE,warning=FALSE}
head(surge_Camille$points)
```
