---
title: "Storm-Surge"
---

Storm surge is arguably the most destructive phenomenon associated with tropical cyclones. Unfortunately, it also difficult to predict or capture, due to the complex interactions between tidal dynamics, wind, rain, bathymetry and topography, and human coastal engineering, among others. The current standard models utilized by the US National Hurricane Center are the Sea Lake and Overland Surges from Hurricanes (SLOSH) models, which incorporate pre-processed 'basins' (currently limited only to the Gulf and Atlantic US coastlines) as well as wind and tidal inputs to solve complex differential equations governing the motion of fluids. Cyclones currently does not do this, but hopes to incorporate something similar in the future. 

For now, data for storm surge is the least complete relative to the other products of wind and precipitation. This is mostly due to the above mentioned complexities and a sparsity of global storm surge data. The Cyclones method involves comparing available water level data with elevation data. This assumes surge waters flow evenly around bathymetric and geographic features, which is overly simplistic. Future methods will attempt to incorporate complex coastal hydrology, such as those implemented in the SLOSH models.

Around the United States and it's territories, and especially for more recent storms, there are farm more water level monitoring locations available for the analysis when compared to other locations and more historic storms. When US monitoring data are unavailable, they are supplemented with the Coupled Model Intercomparison Project Phase 6 (CMIP6) data. Although CMIP6 data is available globally and for more historic storms, these modeled water levels are known to under estimate storm surge, which should be considered when evaluating surge products.

The get_surge() function will provide a raster of maximum surge water levels above a given elevation, as well as a simple spatial features object of the water level observations that were incorporated into the analysis. The datum for flooding is the Mean High High Tide or Mean High High Water (MHHW), so surge values are only returned if water levels are higher than this value at any given point.

Unlike the wind and precipitation methods, get_surge() does not allow for spatial or temporal resolution arguments. This is primarily because the elevation data is already high resolution (30 m), and because water depths are not easily generalized across large areas of different elevation. Thus, the resulting surge product is also high resolution and very expansive when compared to wind and precipitation. For this reason, resulting rasters are relatively large and could easily become obtrusive on even high-performance machines if higher spatial or temporal resolution were requested. For now, get_surge() products are limited to one layer only of the maximum surge depth, and at a moderate spatial resolution of 100 m. 


```{r get_surge_false,eval=FALSE}
###  as with precipitation, surge requires some pre-registration before remote access to the data

###  Elevation comes from the NASA Shuttle Radar Topography Mission
###  it is obtained through the same earthdatalogin methods as the Global Precipitation Mission
###  so if that was already setup, it should be ready
###  if not, see earthdatalogin for permanent registration and setup
###  unlike water levels, elevation data is constant for a given area, regardless of the storm time
###  so, once data for a location has been downloaded (and saved utilizing the dpath arg) it can be
###  directly accessed without downloading for future analyses over the same area

###  likewise, CMIP6 data is obtained through the European Centre for Medium-Range Weather Forecasts (ecmwf)
###  so, if that was already setup for the precipitation data, it will serve here as well
###  if not, see ecmwfr::wf_get_key() and ecmwfr::wf_set_key() for setup
###  for storms after 2024, CMIP6 does not offer daily maximum water levels, only 10 min interval levels
###  so, these downloads will take considerably longer

###  NOAA water levels are retrieved through an anonymous api, so no pre-registration is required
###  more info here: https://api.tidesandcurrents.noaa.gov/api/prod/
###  It should be noted that the server does implement throttling
###  Cyclones methods doe introduce sleep statements to prevent throttling, but users may still encounter issues 
###  when requesting large amounts of data

###  Finally, USGS water level data is also retreived though a semi-anonymous API
###  Users may introduce their own api key, which can be obtained here: https://api.waterdata.usgs.gov/signup/
###  This is not necessary, however, as data will still be retreived without a key
###  but it may boost performance

### as with the other functions, dpath will save component data in a directory for direct future access (no download necessary)
surge_Camille <- get_surge(storms$CAMILLE_1969_NA_1969226N18280,dpath="./Storm Surge/")

ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge level\nabove MHHW\n(m)")+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf_text(data=st_transform(rnaturalearth::ne_countries(type="tiny_countries"),crs(windsSinmsw)),
  #             aes(label = admin), size = 3, check_overlap = TRUE) +
  ggtitle("Hurricane Camille Storm Surge at Louisiana")+
  xlim(-93,-83)+ylim(28,32)+
  theme(axis.title = element_blank(),panel.grid=element_blank())

```

```{r plot_surge,echo=FALSE,message=FALSE,warning=FALSE,fig.width=7,fig.height=10,fig.cap="Fig 6. Results of get_surge() are a single layer only, representing the maximum surge depth above ground for each location during the storm, as well as a corresponding summary table of water level observations. Surge depth is a simple difference between an interpolation of nearby water level observations and the ground elevation. Only depths greater than the mean high high tide are returned. For storms outside of United States jurisdictions, and for any storm before 2000, water level observations are scarce and the CMIP6 model is usually used instead. This model is known to underestimate extreme surge values. Water level sources can be viewed via the corresponding table produced for each storm (below)."}
library(terra)
library(sf)
library(ggplot2)
library(dplyr)
library(patchwork)
windextents <- readRDS("I:/Cyclone Products/Wind/windextents_60.rds")
surge_Camille <- list(rast=rast("I:/Cyclone Products/Storm Surge/v2/CAMILLE_1969_NA_1969226N18280_surge.tif"),points=read.csv("I:/Cyclone Products/Storm Surge/v2/CAMILLE_1969_NA_1969226N18280_surgepoints.csv"))
#surge_Sinlaku <- list(rast=rast("I:/Cyclone Products/Storm Surge/v2/SINLAKU_2026_WP_2026099N09152_surge.tif"),points=read.csv("I:/Cyclone Products/Storm #Surge/v2/SINLAKU_2026_WP_2026099N09152_surgepoints.csv"))
surge_Sandy <- list(rast=rast("I:/Cyclone Products/Storm Surge/v2/SANDY_2012_NA_2012296N14283_surge.tif"),
                    points=read.csv("I:/Cyclone Products/Storm Surge/v2/SANDY_2012_NA_2012296N14283_surgepoints.csv"))

#  Sandy
# ggplot(data=as.data.frame(crop(surge_Sandy$rast,ext(c(-77,-65,34,43))),xy=TRUE)|>rename(value=3)) +
#   geom_raster(aes(x=x,y=y,fill=value))+
#   scale_fill_viridis_c(n=10,na.value = "transparent",name="Flooding Depth (m)")+
#   coord_sf(crs ="+proj=robin +lon_0=180")+
#   theme_bw()+
#   geom_sf(data=st_as_sf(surge_Sandy$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
#  # geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
#   #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
#   ggtitle("Hurricane Camille Storm Surge at Louisiana")+
#   xlim(-77,-65)+ylim(34,43)+
#   theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
# 
# ggplot(data=as.data.frame(crop(surge_Sandy$rast,ext(c(-75,-73,40,42))),xy=TRUE)|>rename(value=3)) +
#   geom_raster(aes(x=x,y=y,fill=value))+
#   scale_fill_viridis_c(n=10,na.value = "transparent",name="Flooding Depth (m)")+
#   coord_sf(crs ="+proj=robin +lon_0=180")+
#   theme_bw()+
#   geom_sf(data=st_as_sf(surge_Sandy$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
#   #geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
#   #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
#   ggtitle("Hurricane Camille Storm Surge at Louisiana")+
#   xlim(-75,-73)+ylim(40,42)+
#   theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
# 
# ggplot(data=as.data.frame(crop(surge_Sandy$rast,ext(c(-74.3,-73,40.4,41))),xy=TRUE)|>rename(value=3)) +
#   geom_raster(aes(x=x,y=y,fill=value))+
#   scale_fill_viridis_c(n=10,na.value = "transparent",name="Flooding Depth (m)")+
#   coord_sf(crs ="+proj=robin +lon_0=180")+
#   theme_bw()+
#   geom_sf(data=st_as_sf(surge_Sandy$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
#   geom_label(data=surge_Sandy$points,aes(x=lon,y=lat,label=round(maxWL,1)),position=position_jitter(width=0,height=1))+
#   #geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
#   #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
#   ggtitle("Hurricane Camille Storm Surge at Louisiana")+
#   xlim(-74.3,-73)+ylim(40.4,41)+
#   theme(axis.title = element_blank(),panel.grid=element_blank())
# 
# ggplot(data=as.data.frame(crop(surge_Sandy$rast,ext(c(-74.3,-74,40.45,40.7))),xy=TRUE)|>rename(value=3)) +
#   geom_raster(aes(x=x,y=y,fill=value))+
#   scale_fill_viridis_c(n=10,na.value = "transparent",name="Flooding Depth (m)")+
#   coord_sf(crs ="+proj=robin +lon_0=180")+
#   theme_bw()+
#   geom_sf(data=st_as_sf(surge_Sandy$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
#   geom_label(data=surge_Sandy$points,aes(x=lon,y=lat,label=round(maxWL,1)))+
#   #geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
#   geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
#   ggtitle("Hurricane Sandy Storm Surge Depth at Staten Island, USA")+
#   xlim(-74.3,-74)+ylim(40.45,40.72)+
#   ggspatial::annotation_scale(location = "tl", width_hint = 0.5)+
#   theme(axis.title = element_blank(),panel.grid=element_blank())




gg1 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge level\nabove MHHW\n(m)")+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  ggtitle("Hurricane Camille Storm Surge at Louisiana")+
  xlim(-93,-83)+ylim(28,32)+
  ggspatial::annotation_scale(location = "tr", width_hint = 0.5)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
gg2 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge depth\n(m)",
                       breaks=c(0,0.5,1,1.5,2))+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  #geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  xlim(-90,-87)+ylim(29.5,31)+
    ggspatial::annotation_scale(location = "br", width_hint = 0.5)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="none")
gg3 <- ggplot(data=as.data.frame(crop(surge_Camille$rast,ext(c(-93,-83,28,32))),xy=TRUE)|>rename(value=3)) +
  geom_raster(aes(x=x,y=y,fill=value))+
  scale_fill_viridis_c(n=10,na.value = "transparent",name="Surge depth\n(m)",
                       breaks=c(0,0.5,1,1.5,2))+
  coord_sf(crs ="+proj=robin +lon_0=180")+
  theme_bw()+
  geom_sf(data=st_as_sf(surge_Camille$points,coords=c("lon","lat"),crs=4326),aes(col=source))+
  geom_sf(data=windextents$CAMILLE_1969_NA_1969226N18280|>filter(location=="track"),lty=2)+
  geom_sf(data=rnaturalearth::ne_countries(scale=10,country="united states of america"),fill=NA)+
  xlim(-89.5,-89)+ylim(29.8,30.2)+
    ggspatial::annotation_scale(location = "tl", width_hint = 0.5)+
  theme(axis.title = element_blank(),panel.grid=element_blank(),legend.position="bottom")
print(gg1/gg2/gg3)
```
```{r surge_table,echo=FALSE,message=FALSE,warning=FALSE}
head(surge_Camille$points)
```
